---
title: "MuM_PreDiCT: PAS"
output: html_notebook
---

last updated: 2023.03.20 (Lisa Kent)

This notebook:
* reads in raw PAS table from database
* links ICD-10 codes in PAS to defined/named mum-predict phenotypes 
* creates instance level wide DF (i.e. one row per individual/admission date with one-hot dummy variable for each phenotype name)

################


```{r}
if(!require(tidyverse)) install.packages("tidyverse")
library(tidyverse)
```

__Lode odbc (Open Database Connectivity) package__
```{r}
library(odbc)
```

__Declare connection to database__
```{r}
con <- dbConnect(odbc::odbc(), "hirusql3", timeout = 10)
```



__Admissions Data__
```{r}
pas <- dbGetQuery(con,"Select * from [HDRNIHBS_E062].[PAS_1]")
```

```{r}
colnames(pas)
```

```{r}
colnames(pas) <- toupper(colnames(pas))
```


```{r}
glimpse(pas)
#there is trailing white space for some variables
```

```{r}
pas <- pas %>% 
  mutate(across(where(is.character), str_trim))

glimpse(pas)
```

```{r}
# re-order data by study_id ascending 
pas<-pas[order(pas$STUDY_ID),]
head(pas, 20)
```

```{r}
n_distinct(pas$STUDY_ID)
n_distinct(pas$STUDY_ID, pas$ADMISSION_DATE)
```



```{r}
pas_icd <- pas %>%
  select(STUDY_ID, ADMISSION_DATE, contains("DIAGNOSIS"))

head(pas_icd, 20)
```

```{r}
#confirming that all diagnosis variables have been used at least once
pas_icd %>% group_by(SECONDARY_DIAGNOSIS_15) %>% count()
```


```{r}
#create long df of ICD10 codes (with study-id and admission date)
pas_icd_long <- pas_icd %>%
  pivot_longer(cols = PRIMARY_DIAGNOSIS_CODE:SECONDARY_DIAGNOSIS_15, names_to = "VAR", values_to = "ICD10_CODE")
head(pas_icd_long, 20)
nrow(pas_icd_long)
```

```{r}
#remove rows where no ICD10 code recorded
pas_icd_long_narm <- pas_icd_long[!is.na(pas_icd_long$ICD10_CODE), ]
head(pas_icd_long_narm, 20)
nrow(pas_icd_long_narm)
```

__READ IN ICD-10 CODE vs HEALTH CONDITION__
```{r}
# Four characters
icd_list_4char <- read_csv("S:/HBS_E062/2. CODE/code lists/ICD10 codes 2023.02.27.csv")
```

```{r}
# Three characters
icd_list_3char <- read_csv("S:/HBS_E062/2. CODE/code lists/ICD10 codes 3char 2023.02.27.csv")
```


```{r}
#sanity check (are all names using same spelling / terminology across 4 and 3 char lists?)
icd_list_4char$PHENOTYPE<-toupper(icd_list_4char$PHENOTYPE)
icd_list_4char<-icd_list_4char[order(icd_list_4char$PHENOTYPE, icd_list_4char$ICD10_CODE),]
icd_list_4char %>% distinct(PHENOTYPE)

icd_list_3char$PHENOTYPE<-toupper(icd_list_3char$PHENOTYPE)
icd_list_3char<-icd_list_3char[order(icd_list_3char$PHENOTYPE, icd_list_3char$ICD10_CODE),]
icd_list_3char %>% distinct(PHENOTYPE)
```


```{r}
# replace certain characters in phenotype name with "_" in preparation for using these to create column names

icd_list_4char$PHENOTYPE <- gsub("\\)", "", icd_list_4char$PHENOTYPE)
icd_list_4char$PHENOTYPE <- gsub(" |&|\\/|\\,|\\(|\\)|\\-", "_", icd_list_4char$PHENOTYPE)
icd_list_4char$PHENOTYPE <- gsub("__|___", "_", icd_list_4char$PHENOTYPE)
icd_list_4char$PHENOTYPE <- paste0("PAS_", icd_list_4char$PHENOTYPE)
icd_list_4char %>% distinct(PHENOTYPE)


icd_list_3char$PHENOTYPE <- gsub("\\)", "", icd_list_3char$PHENOTYPE)
icd_list_3char$PHENOTYPE <- gsub(" |&|\\/|\\,|\\(|\\)|\\-", "_", icd_list_3char$PHENOTYPE)
icd_list_3char$PHENOTYPE <- gsub("__|___", "_", icd_list_3char$PHENOTYPE)
icd_list_3char$PHENOTYPE <- paste0("PAS_", icd_list_3char$PHENOTYPE)
icd_list_3char %>% distinct(PHENOTYPE)

```


```{r}
# some ICD10 codes in PAS are greater than 5 characters (including fullstop)
# e.g. "M24.24" - doesnt exist in the ICD10 browser or mum-predict phenome
# e.g. "A09.9 D" - doesnt exist in the ICD10 browser  or mum-predict phenome - many have this extra " D" at end

# for 4 characters ICD-10 code we extract the first 5 chars (i.e. XXX.X)
pas_icd_long_narm$ICD10_CODE_4CHAR <- substr(pas_icd_long_narm$ICD10_CODE, 1, 5)

# for 3 characters ICD-10 code we extract the first 3 chars (i.e. XXX)
pas_icd_long_narm$ICD10_CODE_3CHAR <- substr(pas_icd_long_narm$ICD10_CODE, 1, 3)
```

```{r}
# for each 4 character ICD10 code in PAS, join to named phenotype from mum-predict phenotype-code list
icd_list_4char <- icd_list_4char %>% rename(PHENO_4CHAR = PHENOTYPE)
pas_icd_long_narm <- left_join(pas_icd_long_narm, icd_list_4char, by=c("ICD10_CODE_4CHAR" = "ICD10_CODE"))
```


```{r}
# for each 3 character ICD10 code in PAS, join to named phenotype from mum-predict phenotype-code list
icd_list_3char <- icd_list_3char %>% rename(PHENO_3CHAR = PHENOTYPE)
pas_icd_long_narm <- left_join(pas_icd_long_narm, icd_list_3char, by=c("ICD10_CODE_3CHAR" = "ICD10_CODE"))
```

```{r}
pas_icd_long_narm <- pas_icd_long_narm %>% mutate(
  PAS_PHENO = if_else(!is.na(PHENO_4CHAR), PHENO_4CHAR, PHENO_3CHAR)
)
```




```{r}
#inspect codes not picked up
pas_icd_long_narm %>% 
  filter(is.na(PAS_PHENO)) %>%
  group_by(ICD10_CODE)%>%
  count()

#count instances
pas_icd_long_narm %>% group_by(PAS_PHENO) %>% count()

#count individuals
pas_icd_long_narm %>% group_by(PAS_PHENO) %>% summarise(COUNT = n_distinct(STUDY_ID))

```






```{r}
#remove rows where no phenotype is recorded
pas_icd_long_narm <- pas_icd_long_narm[!is.na(pas_icd_long_narm$PAS_PHENO), ]
```

```{r}
write_csv(pas_icd_long_narm, file=paste0("S:/HBS_E062/3. DATA/pas_icd_long_narm_", format(lubridate::now(), "%Y.%m.%d_%H.%M"),".csv"))
#below for use during testing phase
#write_csv(pas_icd_long_narm, "S:/HBS_E062/3. DATA/ARCHIVE/pas_icd_long_narm.csv")
```



```{r}
pas_icd_long_narm_summary <- pas_icd_long_narm %>%
  select(STUDY_ID, ADMISSION_DATE, PAS_PHENO) %>%
  group_by(STUDY_ID, ADMISSION_DATE, PAS_PHENO) %>%
  summarise(COUNT = n())
```
```{r}
pas_icd_long_narm_summary <- pas_icd_long_narm_summary %>% ungroup()
```


```{r}
pas_icd_long_narm_summary$COUNT[pas_icd_long_narm_summary$COUNT>0] <- 1
```

#######################################################

__NIMATS dates__
```{r}
#Read in NIMATS dates

nimats_dates <- read_csv("S:/HBS_E062/3. DATA/nimats_dates_2023.06.06_14.15.csv")
```

__Join PAS and NIMATS Dates__
```{r}
pas_dates_long <- left_join(pas_icd_long_narm_summary, nimats_dates, by="STUDY_ID")
```

```{r}
pas_dates_long <- pas_dates_long %>%
  mutate(PAS_PHENO_AGE11 = ifelse(ADMISSION_DATE >= DATE_AGE11, 1, 0),    #age 11+?
         PAS_PHENO_PRE_PREG = ifelse(ADMISSION_DATE <= LMP_CALC, 1, 0)    #pre-pregnancy?
         )
```



```{r}
#some diagnoses can only be retained if they are present at a time point that falls after the mother is aged 11 years
age_11plus <- c("PAS_ATOPIC_ECZEMA", "PAS_ALLERGIC_RHINITIS", "PAS_OTHER_HEADACHES", "PAS_ASTHMA")
pas_dates_long <- pas_dates_long %>%
  mutate(PAS_PHENO_KEEP = if_else(
    PAS_PHENO_PRE_PREG == 1 & (!(PAS_PHENO %in% age_11plus) | (PAS_PHENO %in% age_11plus & PAS_PHENO_AGE11==1)), 1, 0)
  )
 
```


```{r}
#sanity check
pas_dates_long %>%
  filter(PAS_PHENO %in% age_11plus) %>%
  head(100)

pas_dates_long %>%
  filter(PAS_PHENO_AGE11 == 0) %>%
  head(100)
```


```{r}
pas_dates_long <- pas_dates_long %>% filter(PAS_PHENO_KEEP == 1)
```

```{r}
names(pas_dates_long)
pas_dates_long <- pas_dates_long %>% ungroup()
pas_dates_long <- pas_dates_long %>% select(PREG_ID, ADMISSION_DATE, PAS_PHENO, PAS_PHENO_KEEP)
names(pas_dates_long)
```



```{r}
pas_dates_wide <- pas_dates_long %>%
  pivot_wider(names_from = PAS_PHENO,
              values_from = PAS_PHENO_KEEP,
              values_fill = list(PAS_PHENO_KEEP = 0))
```


```{r}
nrow(pas_dates_wide)
pas_dates_wide <- distinct(pas_dates_wide)

nrow(pas_dates_wide)
n_distinct(pas_dates_wide$PREG_ID)
#some pregnancies have more than one row
```

```{r}
#we want to keep any flagged condition within the same pregnancy
#so within each pregnancy we retain the max value (i.e. the 1 one) for any condition
pas_dates_wide[is.na(pas_dates_wide)] <- 0

pas_dates_wide_reduce <- pas_dates_wide %>%
  group_by(PREG_ID) %>%
  summarise(across(starts_with("PAS"), max))

nrow(pas_dates_wide)
n_distinct(pas_dates_wide$PREG_ID)
nrow(pas_dates_wide_reduce)
n_distinct(pas_dates_wide_reduce$PREG_ID)

```
```{r}
#sanity check
pas_dates_wide %>% summarise(across(starts_with("PAS"), sum, na.rm = T))
pas_dates_wide %>% group_by(PAS_ALL_OTHER_CANCERS) %>% summarise(COUNT = n_distinct(PREG_ID))
pas_dates_wide %>% group_by(PAS_CHRONIC_LIVER_DISEASE) %>% summarise(COUNT = n_distinct(PREG_ID))
pas_dates_wide_reduce %>% group_by(PAS_CYSTIC_FIBROSIS) %>% summarise(COUNT = n_distinct(PREG_ID))

pas_dates_wide_reduce %>% summarise(across(starts_with("PAS"), sum, na.rm = T))
pas_dates_wide_reduce %>% group_by(PAS_ALL_OTHER_CANCERS) %>% summarise(COUNT = n_distinct(PREG_ID))
pas_dates_wide_reduce %>% group_by(PAS_CHRONIC_LIVER_DISEASE) %>% summarise(COUNT = n_distinct(PREG_ID))
pas_dates_wide_reduce %>% group_by(PAS_CYSTIC_FIBROSIS) %>% summarise(COUNT = n_distinct(PREG_ID))


```


```{r}

write_csv(pas_dates_wide_reduce, file=paste0("S:/HBS_E062/3. DATA/pas_wide_", format(lubridate::now(), "%Y.%m.%d_%H.%M"),".csv"))

# saving to ARCHIVE while in test phase
#write_csv(pas_dates_wide_reduce, "S:/HBS_E062/3. DATA/ARCHIVE/pas_wide.csv")

```


######################################
#######           END          #######
######################################