---
title: "MuM_PreDiCT: EPD Medications - Physical Health Conditions - Wide"
output: html_notebook
---

last updated: 2023.06.07 (Lisa Kent)

This notebook:
* Operationalises the MuM-PreDiCT phenome definitions related to medications for physical conditions
* Due to the scan rates of prescriptions in NI being less than 100% coverage, several definitions are compared (e.g. 4 vs 3 vs 2 prescriptions in any given 12mo period)




```{r}
if(!require(tidyverse)) install.packages("tidyverse")
library(tidyverse)
```

__READ IN LONG FORMAT PHYSICAL PHENOTYPE__


__READ IN NIMATS DATES__


__JOIN PHYSICAL PHENOTYPE MEDS TO NIMATS DATES__
```{r}
nimats_dates$STUDY_ID <- as.numeric(nimats_dates$STUDY_ID)
epd_pheno_physical_long$STUDY_ID <- as.numeric(epd_pheno_physical_long$STUDY_ID)
nimats_epd_physical <- left_join(nimats_dates, epd_pheno_physical_long, by = "STUDY_ID", multiple = "all", relationship = "many-to-many")
```



__TIMING OF PRESCRIPTIONS__
```{r}
#sanity check
nimats_epd_physical <- nimats_epd_physical %>% 
  arrange(STUDY_ID, PREG_ID, PHENO, ISSUE_DATE_NONMISS)

head(nimats_epd_physical, 100)
```

```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  group_by(PREG_ID, PHENO) %>%
  mutate(DATE_FOURTH_PRESC = lead(ISSUE_DATE_NONMISS,3))
```

```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  group_by(PREG_ID, PHENO) %>%
  mutate(DATE_THIRD_PRESC = lead(ISSUE_DATE_NONMISS,2))
```

```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  group_by(PREG_ID, PHENO) %>%
  mutate(DATE_SECOND_PRESC = lead(ISSUE_DATE_NONMISS,1))
```


```{r}
#nb - this chunk slow to run
library(lubridate)
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(FU_PERIOD_4PRESC = as.numeric(DATE_FOURTH_PRESC - ISSUE_DATE_NONMISS)) %>%
  mutate(ANY_12MO_4PRESC = if_else(FU_PERIOD_4PRESC <=365 & FU_PERIOD_4PRESC >=0, 1, 0))%>%
  
  mutate(FU_PERIOD_3PRESC = as.numeric(DATE_THIRD_PRESC - ISSUE_DATE_NONMISS)) %>%
  mutate(ANY_12MO_3PRESC = if_else(FU_PERIOD_3PRESC <=365 & FU_PERIOD_3PRESC >=0, 1, 0)) %>%
  
  mutate(FU_PERIOD_2PRESC = as.numeric(DATE_SECOND_PRESC - ISSUE_DATE_NONMISS)) %>%
  mutate(ANY_12MO_2PRESC = if_else(FU_PERIOD_2PRESC <=365 & FU_PERIOD_2PRESC >=0, 1, 0)) %>%
  mutate(ANY_6MO_2PRESC = if_else(FU_PERIOD_2PRESC <=183 & FU_PERIOD_2PRESC >=0, 1, 0))

head(nimats_epd_physical, 100)
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(ANY_12MO_4PRESC) %>% count()
nimats_epd_physical %>% group_by(ANY_12MO_3PRESC) %>% count()
nimats_epd_physical %>% group_by(ANY_12MO_3PRESC) %>% count()
nimats_epd_physical %>% group_by(ANY_6MO_2PRESC) %>% count()
```


```{r}
#IDENTIFY IF PRESCRIPTION DISPENSED PRE-PREGNANCY
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(PRE_PREG_DISPENSE = ifelse(ISSUE_DATE_NONMISS <= LMP_CALC, 1, 0))

head(nimats_epd_physical, 100)
```


```{r}
#FLAG IF MEDICATION ISSUED AFTER AGE 11 
#(NB - THIS CRITERIA ONLY APPLICABLE TO SOME PHENOTYPES)
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(PHENO_AGE11 = ifelse(ISSUE_DATE_NONMISS >= DATE_AGE11, 1, 0))

head(nimats_epd_physical, 100)
```


```{r}
library(lubridate)
# FLAG IF ISSUE DATE WITHININ 12MONTHS PRE-PREGNANCY 
#(NB - "ACTIVE" DEFINITION ONLY APPLICABLE TO SOME PHENOTYPES)
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(DATE_12MONTH = ymd(LMP_CALC) - years(1)) %>%
  mutate(PHENO_12MONTH_PRE_PREG = ifelse(ISSUE_DATE_NONMISS >= DATE_12MONTH & ISSUE_DATE_NONMISS <= LMP_CALC, 1, 0))

head(nimats_epd_physical, 100)
```




__TIME DEPENDENT PHYSICAL HEALTH FLAGS__
```{r}
#asthma=> script age11+, with at least 4 scripts in any 12 month period, and at least 1 script in 12months preceding pregnancy
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(EPD_ASTHMA_KEEP_4PRESC = if_else(
    PHENO=="EPD_ASTHMA" & PHENO_AGE11==1 & ANY_12MO_4PRESC==1 & PHENO_12MONTH_PRE_PREG==1,1,0),
    #SCAN RATES OF PRESCRIPTIONS MAY BE LIMITING THE DETECTION RATE OF ASTHMA
    #TO TEST IF DETECTION RATES CHANGE IF WE DROP THE 12 MONTH PRE-PREG CRITERIA
    #asthma=> script age11+, with at least 4 scripts in any 12 month period
    EPD_ASTHMA_KEEP_ALT = if_else(PHENO=="EPD_ASTHMA" & PHENO_AGE11==1 & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #asthma=> script age11+, with at least **3** scripts in any 12 month period, and at least 1 script in 12months preceding pregnancy
    EPD_ASTHMA_KEEP_3PRESC = if_else(PHENO=="EPD_ASTHMA" & PHENO_AGE11==1 & ANY_12MO_3PRESC==1 & PHENO_12MONTH_PRE_PREG==1,1,0),
    
    EPD_ASTHMA_KEEP_2PRESC = if_else(PHENO=="EPD_ASTHMA" & PHENO_AGE11==1 & ANY_12MO_2PRESC==1 & PHENO_12MONTH_PRE_PREG==1,1,0)
    )
 
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_ALT == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_ALT == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ASTHMA_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```


```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    EPD_ATOPIC_ECZEMA_KEEP_4PRESC = if_else(PHENO=="EPD_ATOPIC_ECZEMA" & PHENO_AGE11==1 & ANY_12MO_4PRESC==1 & PHENO_12MONTH_PRE_PREG==1,1,0),
    EPD_ATOPIC_ECZEMA_KEEP_ALT = if_else(PHENO=="EPD_ATOPIC_ECZEMA" & PHENO_AGE11==1 & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    EPD_ATOPIC_ECZEMA_KEEP_3PRESC = if_else(PHENO=="EPD_ATOPIC_ECZEMA" & PHENO_AGE11==1 & ANY_12MO_3PRESC==1 & PHENO_12MONTH_PRE_PREG==1,1,0),
    EPD_ATOPIC_ECZEMA_KEEP_2PRESC = if_else(PHENO=="EPD_ATOPIC_ECZEMA" & PHENO_AGE11==1 & ANY_12MO_2PRESC==1 & PHENO_12MONTH_PRE_PREG==1,1,0)
    
    )
  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_ALT == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_ALT == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ATOPIC_ECZEMA_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_ATOPIC_ECZEMA") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_ATOPIC_ECZEMA") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```


__ALLERGIC_RHINOCONJUNCTIVITIS__
```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 2 prescriptions in any 6 month period, at least one over age 11 years and at least one within 12 mo before preg
    EPD_ALLERGIC_RHINO_KEEP_2PRESC6MO = if_else(PHENO=="EPD_ALLERGIC_RHINO" & PHENO_AGE11==1 & ANY_6MO_2PRESC==1 & PHENO_12MONTH_PRE_PREG==1,1,0),
    #at least 1 prescriptions over age 11 years and at least one within 12 mo before preg
    EPD_ALLERGIC_RHINO_KEEP_ANY = if_else(PHENO=="EPD_ALLERGIC_RHINO" & PHENO_AGE11==1 & PHENO_12MONTH_PRE_PREG==1,1,0),
    #at least 2 prescriptions in any 6 month period, at least one over age 11 years and before pregnancy
    EPD_ALLERGIC_RHINO_KEEP_ALT = if_else(PHENO=="EPD_ALLERGIC_RHINO" & PHENO_AGE11==1 & ANY_6MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    
    )
  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_ALLERGIC_RHINO_KEEP_2PRESC6MO == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ALLERGIC_RHINO_KEEP_2PRESC6MO == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_ALLERGIC_RHINO_KEEP_ANY == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ALLERGIC_RHINO_KEEP_ANY == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_ALLERGIC_RHINO_KEEP_ALT == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_ALLERGIC_RHINO_KEEP_ALT == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_ALLERGIC_RHINO") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_ALLERGIC_RHINO") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```



__HYPERTENSION__
```{r}

nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_HYPERTENSION_KEEP_4PRESC = if_else(PHENO=="EPD_HYPERTENSION" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_HYPERTENSION_KEEP_3PRESC = if_else(PHENO=="EPD_HYPERTENSION" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_HYPERTENSION_KEEP_2PRESC = if_else(PHENO=="EPD_HYPERTENSION" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_HYPERTENSION_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPERTENSION_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_HYPERTENSION_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPERTENSION_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_HYPERTENSION_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPERTENSION_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_HYPERTENSION") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_HYPERTENSION") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```




__HYPERTHYROIDISM__
```{r}

nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_HYPERTHYROIDISM_KEEP_4PRESC = if_else(PHENO=="EPD_HYPERTHYROIDISM" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_HYPERTHYROIDISM_KEEP_3PRESC = if_else(PHENO=="EPD_HYPERTHYROIDISM" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_HYPERTHYROIDISM_KEEP_2PRESC = if_else(PHENO=="EPD_HYPERTHYROIDISM" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_HYPERTHYROIDISM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPERTHYROIDISM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_HYPERTHYROIDISM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPERTHYROIDISM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_HYPERTHYROIDISM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPERTHYROIDISM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_HYPERTHYROIDISM") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_HYPERTHYROIDISM") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```



__HYPOTHYROIDISM__
```{r}

nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_HYPOTHYROIDISM_KEEP_4PRESC = if_else(PHENO=="EPD_HYPOTHYROIDISM" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_HYPOTHYROIDISM_KEEP_3PRESC = if_else(PHENO=="EPD_HYPOTHYROIDISM" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_HYPOTHYROIDISM_KEEP_2PRESC = if_else(PHENO=="EPD_HYPOTHYROIDISM" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_HYPOTHYROIDISM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPOTHYROIDISM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_HYPOTHYROIDISM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPOTHYROIDISM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_HYPOTHYROIDISM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_HYPOTHYROIDISM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_HYPOTHYROIDISM") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_HYPOTHYROIDISM") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```



__IRRITABLE BOWEL DISEASE__
```{r}

nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_IRR_BOWEL_DIS_KEEP_4PRESC = if_else(PHENO=="EPD_IRR_BOWEL_DIS" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_IRR_BOWEL_DIS_KEEP_3PRESC = if_else(PHENO=="EPD_IRR_BOWEL_DIS" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_IRR_BOWEL_DIS_KEEP_2PRESC = if_else(PHENO=="EPD_IRR_BOWEL_DIS" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_IRR_BOWEL_DIS_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_IRR_BOWEL_DIS_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_IRR_BOWEL_DIS_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_IRR_BOWEL_DIS_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_IRR_BOWEL_DIS_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_IRR_BOWEL_DIS_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_IRR_BOWEL_DIS") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_IRR_BOWEL_DIS") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```



__MIGRAINE__
```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_MIGRAINE_KEEP_4PRESC = if_else(PHENO=="EPD_MIGRAINE" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_MIGRAINE_KEEP_3PRESC = if_else(PHENO=="EPD_MIGRAINE" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_MIGRAINE_KEEP_2PRESC = if_else(PHENO=="EPD_MIGRAINE" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_MIGRAINE_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_MIGRAINE_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_MIGRAINE_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_MIGRAINE_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_MIGRAINE_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_MIGRAINE_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_MIGRAINE") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_MIGRAINE") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```



__PSORIASIS__
```{r}
###??? Query = definition of active not clear in phenome word doc.
# do we need to consider topical treatments in the last 12 months? or only psoriasis preparations?

nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_PSORIASIS_KEEP_4PRESC = if_else(PHENO=="EPD_PSORIASIS" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_PSORIASIS_KEEP_3PRESC = if_else(PHENO=="EPD_PSORIASIS" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_PSORIASIS_KEEP_2PRESC = if_else(PHENO=="EPD_PSORIASIS" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_PSORIASIS_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_PSORIASIS_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_PSORIASIS_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_PSORIASIS_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_PSORIASIS_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_PSORIASIS_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_PSORIASIS") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_PSORIASIS") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```


__TYPE 1 DIABETES MELLITUS__
```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_T1DM_KEEP_4PRESC = if_else(PHENO=="EPD_T1DM" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_T1DM_KEEP_3PRESC = if_else(PHENO=="EPD_T1DM" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_T1DM_KEEP_2PRESC = if_else(PHENO=="EPD_T1DM" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_T1DM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_T1DM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_T1DM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_T1DM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_T1DM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_T1DM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_T1DM") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_T1DM") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```


__TYPE 2 DIABETES MELLITUS__
```{r}
nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_T2DM_KEEP_4PRESC = if_else(PHENO=="EPD_T2DM" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_T2DM_KEEP_3PRESC = if_else(PHENO=="EPD_T2DM" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_T2DM_KEEP_2PRESC = if_else(PHENO=="EPD_T2DM" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_T2DM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_T2DM_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_T2DM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_T2DM_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_T2DM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_T2DM_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_T2DM") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_T2DM") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```


__EPILEPSY__
```{r}
###??? Query = Do we need to check for X number of scripts in x months or are we defining epilepsy by at least one prescription at any point in time before pregnancy

nimats_epd_physical <- nimats_epd_physical %>%
  mutate(
    #at least 4 prescriptions in any 12 months
    EPD_EPILEPSY_KEEP_4PRESC = if_else(PHENO=="EPD_EPILEPSY" & ANY_12MO_4PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 3 prescriptions in any 12 months
    EPD_EPILEPSY_KEEP_3PRESC = if_else(PHENO=="EPD_EPILEPSY" & ANY_12MO_3PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    #at least 2 prescriptions in any 12 months
    EPD_EPILEPSY_KEEP_2PRESC = if_else(PHENO=="EPD_EPILEPSY" & ANY_12MO_2PRESC==1 & PRE_PREG_DISPENSE==1,1,0),
    EPD_EPILEPSY_KEEP_ANY = if_else(PHENO=="EPD_EPILEPSY" & PRE_PREG_DISPENSE==1,1,0)
    )
 
head(nimats_epd_physical, 500)  
```


```{r}
#sanity check
nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_4PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_3PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)

nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_2PRESC == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_ANY == 1) %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)
nimats_epd_physical %>% group_by(EPD_EPILEPSY_KEEP_ANY == 1) %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)


nimats_epd_physical %>% group_by(PHENO == "EPD_EPILEPSY") %>% summarise(COUNT = n_distinct(STUDY_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$STUDY_ID)*100)

nimats_epd_physical %>% group_by(PHENO == "EPD_EPILEPSY") %>% summarise(COUNT = n_distinct(PREG_ID),
                                                                  PERC = COUNT/n_distinct(nimats_epd_physical$PREG_ID)*100)
```
###################################


```{r}
nimats_epd_physical_reduce <- nimats_epd_physical %>% select(PREG_ID, starts_with("EPD"))
names(nimats_epd_physical_reduce)
```


```{r}
#nimats_epd_physical_reduce[is.na(nimats_epd_physical_reduce)] <- 0
nimats_epd_physical_reduce <- nimats_epd_physical_reduce %>%
  mutate(across(starts_with("EPD_"), ~ if_else(is.na(.x), 0, .x)))
```


```{r}
nimats_epd_physical_reduce <- distinct(nimats_epd_physical_reduce)
```

```{r}
n_distinct(nimats_epd_physical_reduce$PREG_ID)
nrow(nimats_epd_physical_reduce)
```

```{r}
nimats_epd_physical_reduce <- nimats_epd_physical_reduce %>%
  group_by(PREG_ID) %>%
  summarise(across(starts_with("EPD"), max))
```


```{r}
write_csv(nimats_epd_physical_reduce, file=paste0("S:/HBS_E062/3. DATA/epd_physical_wide_", format(lubridate::now(), "%Y.%m.%d_%H.%M"),".csv"))
```




############################################################################
#######                         REMINDERS/ TO DO                     #######
############################################################################

############################################################################



######################################
#######           END          #######
######################################